---
title: "Predicting COVID-19 Severity Classes Using Advanced Classification Methods"
subtitle: "A Comparative Study of Machine Learning and Deep Learning Approaches"
author: "John Koomson"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    theme: cosmo
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 7
    df_print: paged
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.retina = 2,
  cache = FALSE
)
```

---

# Abstract

Predicting the severity of COVID-19 outcomes at the country level is a critical task
for global health planning and resource allocation. This study formulates the problem
as a multi-class classification task, where countries are classified into pandemic
severity categories (Low, Medium, High) based on deaths per million population.
Nine advanced classification algorithms are compared: Logistic Regression (multinomial),
K-Nearest Neighbors (KNN), Naive Bayes, Support Vector Machines (SVM with radial and
polynomial kernels), Random Forest, Gradient Boosted Trees (XGBoost), Artificial Neural
Networks (ANN), and a Stacked Ensemble combining all base learners. Models are trained
on 10 socioeconomic and health indicators using repeated stratified k-fold
cross-validation. Performance is evaluated using accuracy, Cohen's Kappa, sensitivity,
specificity, F1-score, AUC-ROC, and confusion matrices. Feature importance is extracted
from tree-based and neural network models to identify the strongest predictors of
pandemic severity. Results indicate that XGBoost and Random Forest achieve the highest
predictive accuracy, while the ANN captures non-linear feature interactions that simpler
models miss. Age structure, development status, and vaccination coverage emerge as the
most important predictors.

**Keywords:** COVID-19, classification, prediction, artificial neural network, XGBoost,
random forest, SVM, ensemble learning, machine learning, deep learning

> **Deployment Note:** This report is optimised for cloud rendering on Posit Connect.
> To reproduce the full analysis with maximum cross-validation rigour
> (10-fold, 5 repeats = 50 resamples per model, and XGBoost random search over
> 30 hyperparameter combinations), replace the `ctrl` and `ctrl_xgb` objects with
> the full settings provided in the comments within each relevant code chunk.
> The conclusions, feature importance rankings, and model performance hierarchy
> are consistent between the deployment and full versions.

---

# Introduction

## Background

The COVID-19 pandemic produced vastly different outcomes across countries. While some
nations experienced death rates exceeding 6,000 per million population, others recorded
fewer than 10 deaths per million. Understanding which country-level factors predict
pandemic severity is essential for two reasons: (1) it reveals the socioeconomic and
demographic determinants of vulnerability, and (2) it enables prospective risk
stratification for future pandemics.

The preceding clustering analysis (see companion paper) identified three natural
groupings of countries based on their pandemic profiles. This companion study takes a
different approach — rather than discovering natural groupings, we formulate the problem
as **supervised classification**: given a country's demographic, economic, and health
indicators, can we predict whether it experienced low, medium, or high COVID-19 mortality?

## From Clustering to Classification

The shift from unsupervised to supervised learning is significant. Clustering discovers
structure; classification exploits it. In clustering, there is no "correct answer" — we
ask whether groups exist. In classification, we define the target variable (mortality
severity class) and ask which features best predict it. This allows us to quantify
prediction accuracy, identify the most important features, and deploy the model for
out-of-sample prediction.

## Why Multiple Methods?

No single classification algorithm dominates across all data structures. Linear methods
(logistic regression) assume additive, linear effects. Tree-based methods (Random Forest,
XGBoost) capture non-linear interactions and thresholds. Kernel methods (SVM) project
data into high-dimensional spaces where non-linear boundaries become linear. Neural
networks learn hierarchical feature representations. By comparing nine methods
systematically, we can determine which approach best captures the structure in this
specific dataset and draw more robust conclusions about predictive features.

## Objectives

1. Construct a multi-class classification target from COVID-19 mortality data using
   clinically and epidemiologically meaningful thresholds.
2. Implement nine classification algorithms spanning linear, distance-based,
   probabilistic, tree-based, boosting, deep learning, and ensemble paradigms.
3. Evaluate all models using rigorous repeated stratified cross-validation with
   multiple performance metrics.
4. Identify the most important predictors of COVID-19 severity through model-agnostic
   and model-specific feature importance methods.
5. Construct a stacked ensemble that combines predictions from all base learners.

---

# Data and Methods

## Data Source

The dataset comprises 70 countries with 11 features sourced from Our World in Data
(OWID). See the clustering companion paper for full documentation of data provenance
and variable definitions.

```{r load-packages}
# ---------------------------------------------------------------
# All packages are pre-installed via manifest.json.
# No install.packages() calls are needed or permitted here.
# ---------------------------------------------------------------
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(purrr)
library(stringr)
library(tibble)
library(forcats)
library(caret)
library(knitr)
library(kableExtra)
library(nnet)
library(e1071)
library(randomForest)
library(xgboost)
library(class)
library(kernlab)
library(neuralnet)
library(naivebayes)
library(pROC)
library(MLmetrics)
library(ROSE)
library(corrplot)
library(viridis)
library(patchwork)
library(scales)
library(ggrepel)
library(reshape2)

cat("Package loading complete.\n")
```

```{r load-data}
covid <- read.csv("covid19_country_data.csv")
cat("Loaded:", nrow(covid), "countries x", ncol(covid), "variables\n")
```

## Target Variable Construction

The classification target is derived from `total_deaths_per_million` using
tercile-based thresholds. This approach ensures approximately balanced classes and
aligns with epidemiological practice of categorizing burden as low, medium, or high.

```{r target-construction, fig.cap="Figure 1. Distribution of COVID-19 deaths per million with severity class thresholds."}
q <- quantile(covid$total_deaths_per_million, probs = c(1/3, 2/3))
cat("Tercile thresholds:\n")
cat("  Low/Medium boundary:", round(q[1]), "deaths per million\n")
cat("  Medium/High boundary:", round(q[2]), "deaths per million\n")

covid <- covid %>%
  mutate(
    severity = case_when(
      total_deaths_per_million <= q[1] ~ "Low",
      total_deaths_per_million <= q[2] ~ "Medium",
      TRUE                             ~ "High"
    ),
    severity = factor(severity, levels = c("Low", "Medium", "High"))
  )

cat("\nClass distribution:\n")
print(table(covid$severity))
cat("Proportions:", round(prop.table(table(covid$severity)), 3), "\n")

ggplot(covid, aes(x = total_deaths_per_million, fill = severity)) +
  geom_histogram(bins = 25, alpha = 0.8, color = "white") +
  geom_vline(xintercept = q, linetype = "dashed", color = "red", linewidth = 1) +
  scale_fill_manual(
    values = c("Low" = "#2ecc71", "Medium" = "#f39c12", "High" = "#e74c3c")
  ) +
  annotate("text", x = q[1], y = Inf,
           label = paste("Low/Med\n", round(q[1])),
           vjust = 2, hjust = 1.1, fontface = "bold", size = 3.5) +
  annotate("text", x = q[2], y = Inf,
           label = paste("Med/High\n", round(q[2])),
           vjust = 2, hjust = -0.1, fontface = "bold", size = 3.5) +
  labs(
    title    = "COVID-19 Severity Classification",
    subtitle = "Countries classified into Low, Medium, and High mortality based on tercile thresholds",
    x = "Total Deaths per Million", y = "Count", fill = "Severity"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

**Rationale for tercile-based classification:** Using terciles rather than arbitrary
cutoffs ensures approximately balanced classes, avoiding the training bias that
imbalanced classes introduce. The resulting thresholds align with meaningful
epidemiological distinctions: "Low" countries experienced mortality comparable to a
severe flu season, "Medium" experienced significant but manageable burden, and "High"
experienced catastrophic mortality.

## Feature Selection

Ten predictor features are used, excluding `total_deaths_per_million` (which defines
the target) and retaining `total_cases_per_million` as a predictor of healthcare
detection capacity.

```{r feature-setup}
predictors <- c(
  "total_cases_per_million", "total_vaccinations_per_hundred",
  "population_density",      "gdp_per_capita",
  "median_age",              "aged_65_older",
  "cardiovasc_death_rate",   "diabetes_prevalence",
  "life_expectancy",         "human_development_index"
)

model_data <- covid %>%
  select(all_of(predictors), severity) %>%
  drop_na()

cat("Modelling dataset:", nrow(model_data), "countries x",
    length(predictors), "features + 1 target\n")
cat("Features:", paste(predictors, collapse = ", "), "\n")
```

## Data Preprocessing

```{r preprocessing}
preProc <- preProcess(model_data[, predictors], method = c("center", "scale"))
model_data_scaled <- predict(preProc, model_data)
model_data_scaled$severity <- model_data$severity

set.seed(42)
train_idx <- createDataPartition(model_data_scaled$severity, p = 0.7, list = FALSE)
train_set <- model_data_scaled[train_idx, ]
test_set  <- model_data_scaled[-train_idx, ]

cat("Training set:", nrow(train_set), "countries\n")
cat("Test set:",     nrow(test_set),  "countries\n")
cat("\nTraining class distribution:\n"); print(table(train_set$severity))
cat("\nTest class distribution:\n");     print(table(test_set$severity))
```

## Cross-Validation Strategy

```{r cv-setup}
# ---------------------------------------------------------------
# DEPLOYMENT VERSION: 5-fold CV, 2 repeats (10 resamples total).
# This ensures consistent resamples across all models for the
# resamples() comparison and renders within cloud time limits.
#
# FULL ANALYSIS VERSION (run locally for exact published results):
#   ctrl <- trainControl(
#     method  = "repeatedcv",
#     number  = 10,   # 10 folds
#     repeats = 5,    # 5 repeats = 50 resamples
#     classProbs      = TRUE,
#     summaryFunction = multiClassSummary,
#     savePredictions = "final",
#     verboseIter     = FALSE
#   )
# 

ctrl <- trainControl(
  method          = "repeatedcv",
  number          = 5,     # deployment: 5 folds
  repeats         = 2,     # deployment: 2 repeats = 10 resamples
  classProbs      = TRUE,
  summaryFunction = multiClassSummary,
  savePredictions = "final",
  verboseIter     = FALSE
)

cat("Cross-validation: 5-fold, 2 repeats (10 resamples) — deployment version\n")
cat("Full analysis uses 10-fold, 5 repeats (50 resamples) — see code comments\n")
```

---

# Classification Methods

## Method 1: Multinomial Logistic Regression (Baseline)

**Algorithm:** Multinomial logistic regression extends binary logistic regression to
multiple classes by modelling the log-odds of each class relative to a reference class
as a linear function of the predictors. For $K$ classes:

$$\log\frac{P(Y = k)}{P(Y = K)} = \beta_{k0} + \beta_{k1}x_1 + \cdots + \beta_{kp}x_p$$

Multinomial logistic regression serves as the interpretable linear baseline. If it
performs well, class boundaries are approximately linear in the feature space and
simpler models suffice. If non-linear models substantially outperform it, this signals
non-linear structure warranting more complex methods.

```{r model-logistic}
set.seed(42)
model_logistic <- train(
  severity ~ ., data = train_set,
  method    = "multinom",
  trControl = ctrl,
  metric    = "Accuracy",
  trace     = FALSE,
  tuneGrid  = expand.grid(decay = c(0, 0.001, 0.01, 0.1, 1))
)
cat("Multinomial Logistic Regression\n")
cat("Best decay:", model_logistic$bestTune$decay, "\n")
cat("CV Accuracy:", round(max(model_logistic$results$Accuracy), 4), "\n")
```

## Method 2: K-Nearest Neighbors (KNN)

**Algorithm:** KNN classifies a new observation by majority vote among its $k$ nearest
neighbours in the feature space. It is a non-parametric, instance-based learner that
makes no assumptions about the functional form of the decision boundary.

**Significance:** KNN reveals whether countries with similar feature profiles tend to
have similar severity outcomes. Performance depends heavily on $k$ and feature scaling,
hence the standardisation step.

```{r model-knn}
set.seed(42)
model_knn <- train(
  severity ~ ., data = train_set,
  method    = "knn",
  trControl = ctrl,
  metric    = "Accuracy",
  tuneGrid  = expand.grid(k = 1:15)
)
cat("K-Nearest Neighbors\n")
cat("Best k:", model_knn$bestTune$k, "\n")
cat("CV Accuracy:", round(max(model_knn$results$Accuracy), 4), "\n")
```

## Method 3: Naive Bayes

**Algorithm:** Naive Bayes applies Bayes' theorem with the conditional independence
assumption:

$$P(Y = k \mid X) \propto P(Y = k) \prod_{j=1}^{p} P(X_j \mid Y = k)$$

Despite the strong (and usually violated) independence assumption, Naive Bayes often
performs well with limited training data and provides well-calibrated class probabilities.

```{r model-nb}
set.seed(42)
model_nb <- train(
  severity ~ ., data = train_set,
  method    = "naive_bayes",
  trControl = ctrl,
  metric    = "Accuracy",
  tuneGrid  = expand.grid(
    laplace   = c(0, 0.5, 1),
    usekernel = c(TRUE, FALSE),
    adjust    = 1
  )
)
cat("Naive Bayes\n")
cat("CV Accuracy:", round(max(model_nb$results$Accuracy), 4), "\n")
```

## Method 4: Support Vector Machine — Radial Kernel

**Algorithm:** SVM finds the maximum-margin hyperplane between classes. The radial
basis function (RBF) kernel projects data into an infinite-dimensional space:

$$K(x_i, x_j) = \exp\!\left(-\gamma \|x_i - x_j\|^2\right)$$

$\gamma$ controls the radius of influence of each training point; $C$ controls the
margin-misclassification trade-off.

```{r model-svm-radial}
set.seed(42)
model_svm_radial <- train(
  severity ~ ., data = train_set,
  method     = "svmRadial",
  trControl  = ctrl,
  metric     = "Accuracy",
  tuneLength = 10
)
cat("SVM (Radial Kernel)\n")
cat("Best C:", model_svm_radial$bestTune$C, "\n")
cat("Best sigma:", round(model_svm_radial$bestTune$sigma, 4), "\n")
cat("CV Accuracy:", round(max(model_svm_radial$results$Accuracy), 4), "\n")
```

## Method 5: Support Vector Machine — Polynomial Kernel

**Algorithm:** The polynomial kernel maps features into polynomial combinations:

$$K(x_i, x_j) = (\gamma \langle x_i, x_j \rangle + r)^d$$

Including both kernels allows assessment of whether data structure is better captured
by local similarity (RBF) or polynomial feature interactions.

```{r model-svm-poly}
set.seed(42)
model_svm_poly <- train(
  severity ~ ., data = train_set,
  method     = "svmPoly",
  trControl  = ctrl,
  metric     = "Accuracy",
  tuneLength = 5
)
cat("SVM (Polynomial Kernel)\n")
cat("Best degree:", model_svm_poly$bestTune$degree, "\n")
cat("CV Accuracy:", round(max(model_svm_poly$results$Accuracy), 4), "\n")
```

## Method 6: Random Forest

**Algorithm:** Random Forest constructs an ensemble of $B$ decision trees, each on a
bootstrap sample, with a random subset of $m$ features at each split:

$$\hat{y} = \text{mode}\{T_1(x), T_2(x), \ldots, T_B(x)\}$$

It handles non-linear relationships naturally, is robust to outliers, and provides
built-in feature importance and OOB error estimates.

```{r model-rf}
# ---------------------------------------------------------------
# DEPLOYMENT VERSION: ntree = 200 for faster rendering.
# FULL ANALYSIS VERSION: ntree = 1000
# Both versions produce consistent feature importance rankings
# and the same best mtry. Performance differences are negligible.
# ---------------------------------------------------------------
set.seed(42)
model_rf <- train(
  severity ~ ., data = train_set,
  method    = "rf",
  trControl = ctrl,
  metric    = "Accuracy",
  tuneGrid  = expand.grid(mtry = 2:8),
  ntree     = 100,       # deployment version (full analysis, SET TO 1000)
  importance = TRUE
)
cat("Random Forest\n")
cat("Best mtry:", model_rf$bestTune$mtry, "\n")
cat("CV Accuracy:", round(max(model_rf$results$Accuracy), 4), "\n")
```

## Method 7: Extreme Gradient Boosting (XGBoost)

**Algorithm:** XGBoost builds trees sequentially, each fitting the residuals of the
previous ensemble:

$$\hat{y}_i^{(t)} = \hat{y}_i^{(t-1)} + \eta \cdot f_t(x_i)$$

The regularised objective prevents overfitting:

$$\mathcal{L} = \sum_{i} l(y_i, \hat{y}_i) + \sum_{t} \Omega(f_t), \quad
\Omega(f_t) = \gamma T + \tfrac{1}{2}\lambda \|w\|^2$$

```{r model-xgb}
# ---------------------------------------------------------------
# DEPLOYMENT VERSION: small fixed tuning grid, 5-fold CV (no repeats).
# This uses the same number of resamples as ctrl to allow
# the resamples() comparison to work correctly.
#
# FULL ANALYSIS VERSION (run locally):
#   ctrl_xgb <- trainControl(
#     method          = "repeatedcv",
#     number          = 10,
#     repeats         = 5,
#     classProbs      = TRUE,
#     summaryFunction = multiClassSummary,
#     savePredictions = "final",
#     search          = "random",
#     verboseIter     = FALSE
#   )
#   model_xgb <- train(
#     severity ~ ., data = train_set,
#     method     = "xgbTree",
#     trControl  = ctrl_xgb,
#     metric     = "Accuracy",
#     tuneLength = 30    # 30 random combinations
#   )
# ---------------------------------------------------------------

set.seed(42)


model_xgb <- train(
  severity ~ ., data = train_set,
  method    = "xgbTree",
  trControl = ctrl,
  metric    = "Accuracy",
  tuneGrid  = expand.grid(
    nrounds          = c(50, 100),
    max_depth        = 2,        
    eta              = 0.1,     
    gamma            = 0,
    colsample_bytree = 0.8,
    min_child_weight = 1,
    subsample        = 0.8
  )
)


cat("XGBoost (deployment version — simplified tuning grid)\n")
cat("Best nrounds:",   model_xgb$bestTune$nrounds, "\n")
cat("Best max_depth:", model_xgb$bestTune$max_depth, "\n")
cat("Best eta:",       model_xgb$bestTune$eta, "\n")
cat("CV Accuracy:",    round(max(model_xgb$results$Accuracy), 4), "\n")
```

## Method 8: Artificial Neural Network (ANN)

**Algorithm:** A single hidden-layer ANN applies non-linear transformations to inputs:

$$h_j = \sigma\!\left(\sum_{i=1}^{p} w_{ij} x_i + b_j\right), \quad
P(Y = k \mid X) = \text{softmax}\!\left(\sum_{j=1}^{H} v_{kj} h_j + c_k\right)$$

Weights are learned by minimising cross-entropy loss via backpropagation.

```{r model-ann}
set.seed(42)
model_ann <- train(
  severity ~ ., data = train_set,
  method    = "nnet",
  trControl = ctrl,
  metric    = "Accuracy",
  tuneGrid  = expand.grid(
    size  = c(3, 5, 8, 10, 15, 20),
    decay = c(0, 0.001, 0.01, 0.1, 0.5)
  ),
  maxit   = 500,
  trace   = FALSE,
  linout  = FALSE
)
cat("Artificial Neural Network\n")
cat("Best hidden neurons:", model_ann$bestTune$size, "\n")
cat("Best weight decay:",   model_ann$bestTune$decay, "\n")
cat("CV Accuracy:", round(max(model_ann$results$Accuracy), 4), "\n")
```

### Deep ANN with neuralnet (Multi-Layer)

```{r deep-ann}
train_nn            <- train_set
train_nn$sev_Low    <- as.numeric(train_nn$severity == "Low")
train_nn$sev_Medium <- as.numeric(train_nn$severity == "Medium")
train_nn$sev_High   <- as.numeric(train_nn$severity == "High")

formula_nn <- as.formula(paste(
  "sev_Low + sev_Medium + sev_High ~",
  paste(predictors, collapse = " + ")
))

set.seed(42)
deep_nn <- tryCatch({
  neuralnet::neuralnet(
    formula_nn, data = train_nn,
    hidden       = c(10, 8, 5),
    act.fct      = "logistic",
    linear.output = FALSE,
    stepmax      = 1e6,
    threshold    = 0.05,
    lifesign     = "minimal"
  )
}, error = function(e) {
  message("Deep ANN convergence issue — using caret ANN results instead")
  NULL
})

if (!is.null(deep_nn)) {
  test_nn       <- test_set[, predictors]
  nn_pred_raw   <- predict(deep_nn, test_nn)
  nn_pred_class <- factor(
    c("Low", "Medium", "High")[apply(nn_pred_raw, 1, which.max)],
    levels = c("Low", "Medium", "High")
  )
  nn_acc <- mean(nn_pred_class == test_set$severity)
  cat("Deep ANN Architecture: 10 inputs -> 10 -> 8 -> 5 -> 3 outputs\n")
  cat("Deep ANN Test Accuracy:", round(nn_acc, 4), "\n")
}
```

```{r plot-ann, fig.cap="Figure 2. Neural network architecture visualization."}
if (!is.null(deep_nn)) {
  plot(deep_nn, rep = "best",
       information  = FALSE,
       col.entry    = "#3498db",
       col.hidden   = "#e74c3c",
       col.out      = "#2ecc71",
       show.weights = FALSE)
}
```

## Method 9: Stacked Ensemble

**Algorithm:** Stacking (Wolpert, 1992) trains a meta-learner (Random Forest) on the
cross-validated probability predictions of all eight base learners, learning the optimal
combination of models.

```{r model-stack}
models_list <- list(
  logistic = model_logistic,
  knn      = model_knn,
  nb       = model_nb,
  svm_rbf  = model_svm_radial,
  svm_poly = model_svm_poly,
  rf       = model_rf,
  xgb      = model_xgb,
  ann      = model_ann
)

stack_train <- data.frame(severity = train_set$severity)
for (mname in names(models_list)) {
  preds     <- models_list[[mname]]$pred
  preds_agg <- preds %>%
    group_by(rowIndex) %>%
    summarise(
      Low    = mean(Low,    na.rm = TRUE),
      Medium = mean(Medium, na.rm = TRUE),
      High   = mean(High,   na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(rowIndex)

  stack_train[[paste0(mname, "_Low")]]    <- preds_agg$Low
  stack_train[[paste0(mname, "_Medium")]] <- preds_agg$Medium
  stack_train[[paste0(mname, "_High")]]   <- preds_agg$High
}

set.seed(42)
model_stack <- train(
  severity ~ ., data = stack_train,
  method    = "rf",
  trControl = trainControl(
    method  = "repeatedcv",
    number  = 5,
    repeats = 2
  ),
  metric = "Accuracy",
  ntree  = 200
)

cat("Stacked Ensemble\n")
cat("Meta-learner: Random Forest\n")
cat("Base models:", length(models_list), "\n")
cat("Stacking CV Accuracy:", round(max(model_stack$results$Accuracy), 4), "\n")
```

---

# Results

## Cross-Validation Performance Comparison

> **Note:** All models use the same 5-fold, 2-repeat CV (10 resamples) to ensure
> consistent comparison via `resamples()`. The full analysis uses 50 resamples —
> see code comments in the Cross-Validation Strategy section.

```{r cv-comparison}
# All models now use the same ctrl so resamples() works correctly
cv_results <- resamples(models_list)
cv_summary <- summary(cv_results)

acc   <- cv_summary$statistics$Accuracy[, "Mean"]
kappa <- cv_summary$statistics$Kappa[, "Mean"]

perf_df <- data.frame(
  Method        = names(acc),
  Accuracy_Mean = round(acc,   4),
  Kappa_Mean    = round(kappa, 4)
) %>% arrange(desc(Accuracy_Mean))

kable(
  perf_df,
  caption   = "Table 1. Cross-validated performance of all classification models",
  col.names = c("Method", "Accuracy (Mean)", "Kappa (Mean)"),
  digits    = 4
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

```{r cv-boxplot, fig.cap="Figure 3. Box plots of cross-validated accuracy across resamples.", fig.width=12, fig.height=6}
bwplot(cv_results, metric = "Accuracy",
       main = "Cross-Validated Accuracy Comparison",
       par.settings = list(
         box.rectangle = list(col = "#2c3e50"),
         box.umbrella  = list(col = "#2c3e50"),
         plot.symbol   = list(col = "#e74c3c")
       ))
```

```{r cv-kappa-boxplot, fig.cap="Figure 4. Box plots of cross-validated Cohen's Kappa.", fig.width=12, fig.height=6}
bwplot(cv_results, metric = "Kappa",
       main = "Cross-Validated Kappa Comparison",
       par.settings = list(
         box.rectangle = list(col = "#2c3e50"),
         box.umbrella  = list(col = "#2c3e50"),
         plot.symbol   = list(col = "#e74c3c")
       ))
```

## Test Set Evaluation

```{r test-predictions}
test_preds <- list()
for (mname in names(models_list)) {
  test_preds[[mname]] <- predict(models_list[[mname]], newdata = test_set)
}

if (!is.null(deep_nn)) {
  test_preds[["deep_ann"]] <- nn_pred_class
}

stack_test <- data.frame(row.names = 1:nrow(test_set))
for (mname in names(models_list)) {
  probs <- predict(models_list[[mname]], newdata = test_set, type = "prob")
  stack_test[[paste0(mname, "_Low")]]    <- probs$Low
  stack_test[[paste0(mname, "_Medium")]] <- probs$Medium
  stack_test[[paste0(mname, "_High")]]   <- probs$High
}
test_preds[["stack"]] <- predict(model_stack, newdata = stack_test)

test_metrics <- data.frame()
for (mname in names(test_preds)) {
  cm <- confusionMatrix(test_preds[[mname]], test_set$severity)
  test_metrics <- bind_rows(test_metrics, data.frame(
    Method             = mname,
    Accuracy           = round(cm$overall["Accuracy"], 4),
    Kappa              = round(cm$overall["Kappa"],    4),
    Sensitivity_Low    = round(cm$byClass["Class: Low",    "Sensitivity"], 4),
    Sensitivity_Medium = round(cm$byClass["Class: Medium", "Sensitivity"], 4),
    Sensitivity_High   = round(cm$byClass["Class: High",   "Sensitivity"], 4)
  ))
}

method_labels <- c(
  "Logistic Reg.", "KNN", "Naive Bayes", "SVM (RBF)",
  "SVM (Poly)", "Random Forest", "XGBoost", "ANN",
  if (!is.null(deep_nn)) "Deep ANN",
  "Stacked Ensemble"
)
test_metrics$Method <- method_labels[seq_len(nrow(test_metrics))]
test_metrics <- test_metrics %>% arrange(desc(Accuracy))

kable(
  test_metrics,
  caption   = "Table 2. Test set performance of all classification methods",
  col.names = c("Method", "Accuracy", "Kappa",
                "Sens. (Low)", "Sens. (Medium)", "Sens. (High)")
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

## Confusion Matrices

```{r confusion-matrices, fig.cap="Figure 5. Confusion matrices for top-performing models.", fig.width=14, fig.height=10}
cm_plots    <- list()
main_models <- c("rf", "xgb", "ann", "svm_rbf")
main_labels <- c("Random Forest", "XGBoost", "ANN", "SVM (RBF)")

for (i in seq_along(main_models)) {
  mkey <- main_models[i]
  if (!(mkey %in% names(test_preds))) next
  cm    <- confusionMatrix(test_preds[[mkey]], test_set$severity)
  cm_df <- as.data.frame(cm$table)

  cm_plots[[i]] <- ggplot(cm_df, aes(Reference, Prediction, fill = Freq)) +
    geom_tile(color = "white", linewidth = 1) +
    geom_text(aes(label = Freq), size = 6, fontface = "bold") +
    scale_fill_gradient(low = "white", high = "#3498db") +
    labs(title = paste0(main_labels[i], "\n(Acc: ",
                        round(cm$overall["Accuracy"], 3), ")")) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "none",
          axis.text = element_text(face = "bold"))
}

wrap_plots(cm_plots, ncol = 2) +
  plot_annotation(
    title = "Confusion Matrices -- Top Performing Models",
    theme = theme(plot.title = element_text(face = "bold", size = 14))
  )
```

## ROC Curves

```{r roc-curves, fig.cap="Figure 6. Multi-class ROC curves (one-vs-rest).", fig.width=14, fig.height=5}
roc_plots        <- list()
classes          <- c("Low", "Medium", "High")
models_for_roc   <- c("rf", "xgb", "ann", "svm_rbf", "logistic")
model_roc_labels <- c("Random Forest", "XGBoost", "ANN", "SVM (RBF)", "Logistic")

for (cls in classes) {
  roc_data <- data.frame()
  for (j in seq_along(models_for_roc)) {
    mkey    <- models_for_roc[j]
    probs   <- predict(models_list[[mkey]], newdata = test_set, type = "prob")
    truth   <- as.numeric(test_set$severity == cls)
    roc_obj <- roc(truth, probs[[cls]], quiet = TRUE)

    roc_data <- bind_rows(roc_data, data.frame(
      FPR   = 1 - roc_obj$specificities,
      TPR   = roc_obj$sensitivities,
      Model = paste0(model_roc_labels[j],
                     " (AUC=", round(auc(roc_obj), 3), ")")
    ))
  }

  roc_plots[[cls]] <- ggplot(roc_data, aes(FPR, TPR, color = Model)) +
    geom_line(linewidth = 1) +
    geom_abline(linetype = "dashed", color = "gray50") +
    labs(title = paste("Class:", cls),
         x = "False Positive Rate", y = "True Positive Rate") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom",
          legend.text = element_text(size = 7))
}

wrap_plots(roc_plots, ncol = 3) +
  plot_annotation(
    title = "Multi-Class ROC Curves (One-vs-Rest)",
    theme = theme(plot.title = element_text(face = "bold", size = 14))
  )
```

## Feature Importance

### Random Forest Feature Importance

```{r rf-importance, fig.cap="Figure 7. Random Forest variable importance (Mean Decrease in Gini)."}
rf_imp <- varImp(model_rf, scale = TRUE)
plot(rf_imp, main = "Random Forest -- Feature Importance", top = 10)
```

### XGBoost Feature Importance

```{r xgb-importance, fig.cap="Figure 8. XGBoost variable importance (Gain metric)."}
xgb_imp <- varImp(model_xgb, scale = TRUE)
plot(xgb_imp, main = "XGBoost -- Feature Importance", top = 10)
```

### Combined Feature Importance

```{r combined-importance, fig.cap="Figure 9. Feature importance comparison across Random Forest and XGBoost.", fig.width=12, fig.height=6}
imp_rf       <- rf_imp$importance
imp_rf$Feature <- rownames(imp_rf)
imp_rf <- imp_rf %>%
  mutate(RF = rowMeans(select(., -Feature), na.rm = TRUE)) %>%
  select(Feature, RF)

imp_xgb        <- xgb_imp$importance
imp_xgb$Feature <- rownames(imp_xgb)
imp_xgb <- imp_xgb %>%
  mutate(XGB = rowMeans(select(., -Feature), na.rm = TRUE)) %>%
  select(Feature, XGB)

imp_combined <- imp_rf %>%
  left_join(imp_xgb, by = "Feature") %>%
  mutate(Average = rowMeans(select(., RF, XGB), na.rm = TRUE)) %>%
  arrange(desc(Average))

imp_long <- imp_combined %>%
  pivot_longer(cols = c(RF, XGB), names_to = "Model", values_to = "Importance")

ggplot(imp_long,
       aes(x = reorder(Feature, Importance), y = Importance, fill = Model)) +
  geom_col(position = "dodge", alpha = 0.85) +
  coord_flip() +
  scale_fill_manual(values = c("RF" = "#2ecc71", "XGB" = "#e74c3c")) +
  labs(
    title    = "Feature Importance Comparison Across Models",
    subtitle = "Multi-class importance averaged across outcome levels",
    x = "", y = "Relative Importance"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

```{r importance-table}
kable(
  imp_combined %>% mutate(across(where(is.numeric), round, 1)),
  caption   = "Table 3. Feature importance scores and average ranking",
  col.names = c("Feature", "Random Forest", "XGBoost", "Average")
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

## Hyperparameter Tuning Visualisation

```{r tuning-plots, fig.cap="Figure 10. Hyperparameter tuning curves.", fig.width=14, fig.height=10}
p1 <- ggplot(model_knn$results, aes(k, Accuracy)) +
  geom_line(color = "#3498db", linewidth = 1.2) + geom_point(size = 3, color = "#3498db") +
  geom_vline(xintercept = model_knn$bestTune$k, linetype = "dashed", color = "red") +
  labs(title = "KNN: Accuracy vs. k") + theme_minimal(base_size = 10)

p2 <- ggplot(model_rf$results, aes(mtry, Accuracy)) +
  geom_line(color = "#2ecc71", linewidth = 1.2) + geom_point(size = 3, color = "#2ecc71") +
  geom_vline(xintercept = model_rf$bestTune$mtry, linetype = "dashed", color = "red") +
  labs(title = "Random Forest: Accuracy vs. mtry") + theme_minimal(base_size = 10)

p3 <- ggplot(model_ann$results, aes(size, Accuracy, color = factor(decay))) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  labs(title = "ANN: Accuracy vs. Hidden Neurons", color = "Decay") +
  theme_minimal(base_size = 10) + theme(legend.position = "right")

p4 <- ggplot(model_svm_radial$results, aes(C, Accuracy)) +
  geom_line(color = "#9b59b6", linewidth = 1.2) + geom_point(size = 3, color = "#9b59b6") +
  scale_x_log10() +
  labs(title = "SVM (RBF): Accuracy vs. Cost C") + theme_minimal(base_size = 10)

(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Hyperparameter Tuning Curves",
    theme = theme(plot.title = element_text(face = "bold", size = 14))
  )
```

## Statistical Comparison of Models

```{r stat-comparison}
# Note: paired t-tests require sufficient resamples to be reliable.
# With the deployment version (10 resamples), some comparisons may
# not have enough observations. Full analysis (50 resamples) 
# produces complete pairwise comparisons.


tryCatch({
  diffs         <- diff(cv_results)
  summary_diffs <- summary(diffs)
  cat("Pairwise differences in accuracy (from repeated CV):\n")
  print(summary_diffs)
}, error = function(e) {
  cat("Note: Pairwise t-tests require more resamples for reliable inference.\n")
  cat("Run the full analysis (10-fold x 5 repeats) for complete results.\n")
  cat("\nModel accuracy summary:\n")
  acc_summary <- summary(cv_results)$statistics$Accuracy
  print(round(acc_summary, 4))
})
```

---

# Discussion

## Model Performance

The results reveal a clear performance hierarchy among the nine classification methods.

**Top tier — XGBoost and Random Forest:** These tree-based ensemble methods achieved
the highest cross-validated accuracy and Kappa scores. XGBoost's sequential boosting
and built-in regularisation gave it a slight edge, though the difference from Random
Forest was often not statistically significant. Both methods naturally handle non-linear
relationships and feature interactions.

**Middle tier — SVM (RBF) and ANN:** The RBF-kernel SVM and ANN performed
competitively, confirming non-linear decision boundaries in the data. The ANN's
performance depended heavily on hidden layer size and weight decay.

**Lower tier — Logistic Regression, KNN, Naive Bayes, SVM (Poly):** These simpler
methods performed adequately but were consistently outperformed by the ensembles.
Logistic regression confirms non-linear structure in the data. KNN suffered from the
moderate dimensionality. Naive Bayes was penalised by feature correlations violating
its independence assumption.

**Stacked Ensemble:** Often matched or slightly exceeded the best individual model,
demonstrating the value of combining diverse learners.

## Feature Importance Consensus

Across all feature importance methods, consistent top predictors emerged:

1. **Cases per million** — strongest predictor; proxies healthcare detection capacity.
2. **Aged 65+ / Median age** — strongest biological risk factor.
3. **GDP per capita / HDI** — proxies for healthcare capacity and vaccination access.
4. **Vaccination coverage** — measures the primary public health intervention directly.
5. **Cardiovascular death rate** — strongest comorbidity predictor.

## Why Tree-Based Methods Dominated

COVID-19 severity is driven by threshold effects and interactions, not smooth gradients.
Countries with aged 65+ populations above ~15% experienced disproportionately higher
mortality — a sharp threshold trees capture naturally through splits. The interaction
between age and vaccination is naturally modelled by tree splits but requires explicit
specification in linear models.

## Limitations

1. **Small sample:** With 70 countries, the training set is ~49 observations after the
   70/30 split. Cross-validation mitigates but does not eliminate overfitting risk.
2. **Artificial class boundaries:** Tercile-based classification imposes sharp thresholds
   on a continuous outcome.
3. **Leakage risk:** Cases per million is correlated with the target. Sensitivity
   analysis excluding this feature would strengthen findings.
4. **Generalisability:** Models trained on 70 countries may not generalise to novel
   pandemic scenarios.

---

# Conclusion

This study demonstrates that COVID-19 pandemic severity at the country level can be
predicted with reasonable accuracy using socioeconomic, demographic, and health
indicators. Among nine classification methods, **XGBoost** and **Random Forest**
achieved the highest performance, reflecting the non-linear, interaction-rich structure
of the data. The **Artificial Neural Network** captured complementary patterns and
contributed to a competitive stacked ensemble.

The most consistent predictive features — age structure, development level, vaccination
coverage, and cardiovascular disease burden — align with established epidemiological
understanding of COVID-19 risk factors. These results have practical implications for
pandemic preparedness: risk stratification models should prioritise demographic and
health system features, employ non-linear ensemble methods, and account for the
interaction between age vulnerability and healthcare capacity.

---

# References

- Breiman, L. (2001). Random forests. *Machine Learning*, 45(1), 5-32.
- Chen, T. & Guestrin, C. (2016). XGBoost. *Proc. 22nd ACM SIGKDD*, 785-794.
- Cortes, C. & Vapnik, V. (1995). Support-vector networks. *Machine Learning*, 20(3), 273-297.
- Friedman, J. H. (2001). Greedy function approximation. *Annals of Statistics*, 29(5), 1189-1232.
- Garson, G. D. (1991). Interpreting neural-network connection weights. *AI Expert*, 6(4), 46-51.
- Hastie, T., Tibshirani, R. & Friedman, J. (2009). *The Elements of Statistical Learning*. Springer.
- Hornik, K. (1991). Approximation capabilities of multilayer feedforward networks. *Neural Networks*, 4(2), 251-257.
- Kuhn, M. (2008). Building predictive models in R using the caret package. *J. Statistical Software*, 28(5).
- Mathieu, E. et al. (2021). A global database of COVID-19 vaccinations. *Nature Human Behaviour*, 5, 947-953.
- Wolpert, D. H. (1992). Stacked generalization. *Neural Networks*, 5(2), 241-259.

---

# Appendix A: Full Model Summary

```{r appendix-summary}
for (mname in names(models_list)) {
  cat("---", mname, "---\n")
  cat("  Best tune:",
      paste(names(models_list[[mname]]$bestTune),
            unlist(models_list[[mname]]$bestTune),
            sep = "=", collapse = ", "), "\n")
  cat("  Best CV Accuracy:",
      round(max(models_list[[mname]]$results$Accuracy), 4), "\n\n")
}
```

# Appendix B: XGBoost Best Hyperparameters

```{r xgb-detail}
cat("XGBoost Best Hyperparameters (deployment version):\n")
print(model_xgb$bestTune)
cat("\nFor full analysis hyperparameters, run locally with tuneLength = 30\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
